### 2025-10-26 thesis/implementation/index.tex
Add more impl content for the vector db impl

\paragraph{Largest Group Selection Heuristic}

The heuristic method employed by the voting algorithm is to steer the pipeline
to agree with itself – if multiple generation attempts yielded the same result,
these generation attempts are more likely than outliers.

@@

The result-based partitioning approach has limitations with subtle errors that
differ by few rows, as these would be treated as completely different result
groups. Furthermore executing every query candidate against the real database
comes along with a significant performance penalty for expensive queries where
the voting step might incur load onto the database, use significant amounts of
memory for loading in all data into memory and slow down the voting as the
entire result needs to be hashed. This system lacks model uncertainty
calibration for SQL task confidence scoring which could be used as one
dimension for partitioning.

@@

>>>>

\paragraph{Bucket Selection Heuristic}

The heuristic method employed by the voting algorithm is steering the pipeline
to agree with itself – if multiple generation attempts yielded the same result,
these generation attempts are more likely than outliers.

@@

The result-based partitioning approach has limitations with regards to subtle
errors that differ by few rows, as these would be treated as completely
different result buckets. Furthermore executing every query candidate against
the real database comes along with a significant performance penalty for
expensive queries where the voting step might incur load onto the database, use
significant amounts of memory for loading in all data into memory and slow down
the voting as the entire result needs to be hashed. This system lacks model
uncertainty calibration for SQL confidence scoring which could be used as
another dimension for partitioning.

@@

\subsection{Supporting Systems and Optimizations}

The supporting systems of \textsc{Natural} are important tools to simplify the 
actual pipeline development, employing performance optimizations and enable
rapid development of new approaches and hypotheses. The primary support systems
are \texttt{Vector} – a vector database abstraction on top of SQLite,
\texttt{wwl} – a library that implements Wasserstein-Weisfeiler-Leman kernels
\citep{WWL} and the \texttt{embedding} and \texttt{masking} modules to compare
and semantically search the embedding space.

\subsubsection{Vector Database}

The vector database implementation \texttt{Vector} is used both during
sampling to construct the vector db and runtime to run the sample selection
algorithm. Using SQLite and \texttt{sqlite-vec} to implement cosine search in
text corpora of up to a few million samples dramatically simplified the
development of \textsc{Natural} as no manual performance optimizations had to
be implemnented. SQLite has a second architectural benefit besides relatively
good performance, which is the portability characteristics of having all
samples indexed in a single file on disk. This enables swapping out the
respectively used samples by using another \texttt{.vector} file on disk.

\begin{verbatim}
let question = "How many ..";

// Initialize vector database for similarity search
let vector = Vector::infer();
let model = EmbeddingModel::from_env().expect("failed to load embedding model");

// Query distance indices
let index = DistanceIndexRow::get("test_db", &vector)?;

// Embed question and perform similarity search
let selector = Selector::new(&question, None, &model, &vector)?;
let samples = selection::<4>(
    &Selector::new(
        question.as_ref(),
        None,
        &model,
        &vector,
    )?,
    &index.into(),
    &vector,
)?;

\end{verbatim}

\paragraph{Database Schema}

As the \texttt{.vector} file is still a regular SQLite database, alongside
indexing tables using embeddings (\texttt{float[4096]} or \texttt{float[2048]})
regular database tables can be maintained to save data derived at sampling
time. \texttt{Vector} hosts the following database schema:

\begin{verbatim}
CREATE TABLE IF NOT EXISTS whitelist (
  word             text not null primary key,
  frequency        integer not null,
  domain_count     integer not null
);

CREATE TABLE IF NOT EXISTS databases (
  id               text not null,
  name             text not null,
  graph            text not null
);

CREATE TABLE IF NOT EXISTS distance_indices (
  db_id            text not null,
  distances        text not null
);

CREATE VIRTUAL TABLE IF NOT EXISTS samples USING vec0 (
  id                 text not null,
  db_id              text not null,

  question           text not null,
  question_masked    text not null,
  question_embedding float[2048],

  sql                text not null,
  sql_masked         text not null,
  sql_embedding      float[2048]
);
\end{verbatim}

Which maintains the word whitelist for query masking, the database graphs of
the sample databases and the precomputed database indices. Thus at pipeline
runtime the database indices and whitelists can be reconstructed through a
cheap selection from \texttt{Vector} even though they are not indexed through
embeddings.

The schema of the \texttt{samples} table largely drives the capabilities for
example selection at runtime, it maintains embeddings of the masked sql query
and natural language question respectively. Thus after cosine similarity search
the original query or question can be reconstructed as well as the graph layout
of the underlying database can be accessed through the \texttt{database} table.

The \texttt{database} and \texttt{database\_indices} table contain \texttt{JSON}
columns respectively for serializing complex graph and graph index structures
to the database.
